\bf Properties and Bayes fitting of restricted Boltzmann machines

### 2017-07-11 writing/paper-rbm/docs/paper.tex
edits from Dr. Vardeman

While the computational concerns are inconvenient in fitting, issues due
to model flexibility are potentially more concerning. For a RBM model
with enough hidden variables, it has been shown that any distribution
for the visibles can be approximated arbitrarily well (Le Roux and
Bengio \protect\hyperlink{ref-le2008representational}{2008}; Montufar
and Ay \protect\hyperlink{ref-montufar2011refinements}{2011}; and
Montúfar, Rauh, and Ay
\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for cell
data, the empirical distribution of an observed training set of visible
variables provides a highest likelihood benchmark, before any parametric
model class is even introduced and applied to obtain a refinement of
model fit. As a consequence, we find that any fully principled fitting
method based on the likelihood for a RBM with enough hidden variables
will seek to reproduce the (discrete) empirical distribution of a
training set. This aspect can be undesirable, and perhaps even
unexpected compared to most modeling scenarios, in that no ``smoothed
distribution'' may result when fitting a RBM model of sufficient size
with a rigorous likelihood-based method. We are therefore led to be
skeptical that models that involve these structures (like a deep
Boltzmann machine) can achieve useful prediction or inference in a
principled way without intentionally limiting the flexibility of the
fitted model.

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
However, the degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the accepted fitted methodology (Hinton
\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
paper, we attempt to address the concerns with overfitting and poor
mixing in an alternative, and perhaps more transparent manner, via
specification of a Bayesian prior in Section
\ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. We examine three Bayesian fitting techniques intended to
avoid model impropriety in Section \ref{model-fitting} and conclude with
a discussion in Section \ref{discussion}. Appendix \ref{appendix-rbm}
provides proofs for results on RBM parameterizations and data codings
described in Section \ref{data-encoding}.

While applications of the RBM have claimed some successes in
classification problems, it is unclear if the model class allows one to
go beyond fitting to other statistical matters of importance in using
the models, such as quantification of the uncertainty of estimation and
the formulation of predictive distributions. These are closely tied to
the probability properties of the RBM model for explaining data
generation and it is hoped that the current exposition contributes to a
better understanding in this direction.

@@

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}; R. Salakhutdinov and
Hinton \protect\hyperlink{ref-salakhutdinov2012efficient}{2012};
Srivastava, Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
stacking multiple layers of RBMs in a deep architecture, proponents of
the models claim to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(R. Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
is achieved by treating a hidden layer of one RBM as the visible layer
in a second RBM, and so on, until the desired multi-layer architecture
is created.

@@

The highly flexible nature of a RBM (having as it does
\(\nh + \nv + \nh*\nv\) parameters) creates at least three kinds of
potential issues in model impropriety that we will call
\emph{degeneracy}, \emph{instability}, and \emph{uninterpretability}. In
this section we define these characteristics, consider how to quantify
these in a RBM, and point out relationships among them.

@@

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
models, \(\mathcal{X}\) denotes all possible graphs that can be
constructed from a set of nodes and an exponentially parameterized
random graph model has a distribution of the form
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where \(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the
model parameter, and
\(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\) is a vector of
statistics based on the adjacency matrix of a graph. Here, as earlier,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
exponentially parametrized random graph model at \(\boldsymbol \theta\)
as \emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

@@

Considering exponential families of distributions, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) introduced
a concept of model deficiency related to \emph{instability}. Instability
can be roughly thought of as excessive sensitivity in the model, where
small changes in the components of potential data outcomes,
\(\boldsymbol x\), may lead to substantial changes in the probability
function \(f_{\boldsymbol \theta}(\boldsymbol x)\). Furthermore, model
instability can be viewed on a spectrum of potential sensitivity in
probability structure, with model degeneracy included as a extreme or
limiting case of instability. To quantify ``instability'' more
rigorously (particularly beyond the definition given by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011})) it is
useful to consider how RBM models might be expanded to incorporate more
and more visibles. When increasing the size of RBM models, it becomes
necessary to grow the number of model parameters (and in this process
one may also arbitrarily expand the number of hidden variables used). To
this end, let
\(\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1\),
denote an element of a sequence of RBM parameters indexed by the number
\(\nv\) of visibles \((V_1, \dots, V_\nv)\) and define a (scaled)
extremal log-probability ratio of the RBM model at
\(\boldsymbol \theta_{\nv}\) as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eq:elpr}
\end{align}
where
\(P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)\)
is the RBM probability of observing outcome \((v_1, \dots, v_\nv)\) for
the visible variables \((V_1, \dots, V_\nv)\) under parameter vector
\(\boldsymbol \theta_\nv\), after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles
(i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
(\eqref{eq:elpr}) needs to stay bounded for the sequence of RBM models to
be stable. That is, we make the following convention.
\BeginKnitrBlock{definition}[S-unstable RBM]
\protect\hypertarget{def:unnamed-chunk-1}{}{\label{def:unnamed-chunk-1}
\iffalse (S-unstable RBM) \fi{} }Let
\(\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh + \nh*\nv}, \nv \ge 1\),
be an element of a sequence of RBM parameters where the number of
hiddens, \(\nh \equiv \nh(\nv) \ge 1\), can be an arbitrary function of
the number \(\nv\) of visibles. A RBM model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, the RBM model sequence is
unstable if, given any \(c > 0\), there exists an integer \(n_c > 0\) so
that \(\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c\) for all
\(\nv \ge n_c\). This definition of \emph{S-unstable} is a
generalization or re-interpretation of the ``unstable'' concept of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) in that
here RBM models for visibles \((v_1, \dots, v_\nv)\) do not form an
exponential family and the dimensionality of \(\boldsymbol \theta_\nv\)
is not fixed, but rather grows with \(\nv\).

@@

In other words, if the probability ratio (\eqref{eq:elpr}) is too large,
then a RBM model sequence will exhibit large probability shifts for very
small changes in data configurations (i.e., will exhibit instability).
Such instability can be a concern in a model for the similar reasons to
degeneracy: as outcomes vary over the sample space, the geography of
probabilities is extremely rugged, with deep pits following sharp
mountains. Recall the applied example of RBM models as a means to
classify images. For data as pixels in an image, the instability result
in Proposition \ref{prp:instab1} manifests itself as a one pixel change
in an image (one component of the visible vector) resulting in a large
shift in the probability, which in turn could result in a vastly
different classification of the image. Examples of this behavior have
been presented in Szegedy et al.
(\protect\hyperlink{ref-szegedy2013intriguing}{2013}) for other deep
learning models, in which a one pixel change in a test image results in
a wildly different classification.

@@

We next explore and numerically explain the relationship between values
of \(\boldsymbol \theta\) and the three notions of model impropriety,
near-degeneracy, instability, and uninterpretability, for RBM models of
varying sizes.

@@

The results of this numerical study are summarized in Figures
\ref{fig:degen-plots}, \ref{fig:instab-plots}, and
\ref{fig:uninterp-plots}. From these three figures, it is clear that all
three measures of model impropriety show higher values for larger
magnitudes of the parameter vectors, supporting the RBM model properties
developed in Section \ref{background}. As a compounding issue, these
figures show that, as model grow in size, it becomes easier for more
parameter configurations to push RBM models into near-degeneracy,
instability and uninterpretability. Additionally, since there are
\(\nh*\nv\) interaction terms in \(\boldsymbol \theta\) versus only
\(\nh + \nv\) main effect terms, for large models there are many more
interaction parameters than main effects in the models. And so, severely
limiting the magnitude of the individual interactions may well help
prevent model impropriety. \par
\begin{figure}
\includegraphics{paper_files/figure-latex/degen-plots-1} \caption{Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size, where $\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})$ is split into main and interaction parameters. Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.}\label{fig:degen-plots}
\end{figure}
\begin{figure}
\includegraphics{paper_files/figure-latex/instab-plots-1} \caption{The sample mean value of $\text{ELPR}(\boldsymbol \theta)/V$ at each grid point for each combination of magnitude of $\boldsymbol \theta$ and model size. As the magnitude of $\boldsymbol \theta$ grows, so does the value of this metric, indicating typical instability in the model.}\label{fig:instab-plots}
\end{figure}
\begin{figure}
\includegraphics{paper_files/figure-latex/uninterp-plots-1} \caption{The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and the expectation vector given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$. Larger magnitudes of $\boldsymbol \theta$ correspond to larger differences, thus indicating reduced interpretability.}\label{fig:uninterp-plots}
\end{figure}
Figure \ref{fig:split-plots-dist} shows the fraction of near-degenerate
models for each magnitude of \(\boldsymbol \theta\) for each model
architecture. For each number \(\nv\) of visibles in the model, as the
number \(\nh\) of hiddens increase, the fraction near-degenerate
diverges from zero at increasing rates for larger values of
\(||\boldsymbol \theta||\). This shows that, as model size gets larger,
the risk of degeneracy starts at a slightly larger magnitude of
parameters, but very quickly increases until reaching close to 1. \par
\begin{figure}
\includegraphics{paper_files/figure-latex/split-plots-dist-1} \caption{The fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$. For each number $\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\boldsymbol \theta||$ as the number $\nh$ of hidden variables increases and the slope becomes steeper as $\nh$ increases as well.}\label{fig:split-plots-dist}
\end{figure}
These manageable examples indicate that RBMs are near-degenerate,
unstable, and uninterpretable for large portions of the parameter space
with large \(\|\boldsymbol \theta\|\). These problematic aspects require
serious consideration when using RBM models, on top of the additional
matter of principled/rigorous fitting of RBM models.

@@

Typically, fitting a RBM via maximum likelihood (ML) methods will be
infeasible due mainly to the intractability of the normalizing term
\(\gamma(\boldsymbol \theta)\) in a model (\eqref{eq:pmf}) of any
realistic size. Ad hoc methods are used instead, which aim to avoid this
problem by using stochastic ML approximations that employ a small number
of MCMC draws (i.e., contrastive divergence, (Hinton
\protect\hyperlink{ref-hinton2002training}{2002})).

@@

appendix shows, only it is possible to approximate any distribution on
the visibles arbitrarily well (cf. Montufar and Ay

@@

require at least \(4.7\) times as many iterations to be run of the MCMC
chain in order to achieve the same amount of effective information about
the posterior distribution. For this reason, consistent with the ACF
results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective
method for fitting the RBM, though computing resources can hinder use of
the alternative BwTNML involving marginalization of hidden variables.

>>>>

While the computational concerns are inconvenient in fitting, issues due
to model flexibility are potentially more concerning. For a RBM model
with enough hidden variables, it has been shown that any distribution
for the visibles can be approximated arbitrarily well (Le Roux and
Bengio \protect\hyperlink{ref-le2008representational}{2008}; Montufar
and Ay \protect\hyperlink{ref-montufar2011refinements}{2011}; and
Montúfar, Rauh, and Ay
\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for
binary data, the empirical distribution of an observed training set of
visible variables provides a highest likelihood benchmark, before any
parametric model class is even introduced and applied to obtain a
refinement of model fit. As a consequence, we find that any fully
principled fitting method based on the likelihood for a RBM with enough
hidden variables will seek to reproduce the (discrete) empirical
distribution of a training set. This aspect can be undesirable, and
perhaps even unexpected compared to most modeling scenarios, in that no
``smoothed distribution'' may result when fitting a RBM model of
sufficient size with a rigorous likelihood-based method. We are
therefore led to be skeptical that models that involve these structures
(like deep Boltzmann machines) can achieve useful prediction or
inference in a principled way without intentionally limiting the
flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
However, the degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the accepted fitting methodology (Hinton
\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
paper, we attempt to address the concerns with overfitting and poor
mixing in an alternative, and perhaps more transparent manner, via
specification of a Bayesian prior in Section
\ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. We examine three Bayesian fitting techniques intended to
avoid model impropriety issues raised in Section \ref{model-fitting} and
conclude with a discussion in Section \ref{discussion}. Appendix
\ref{appendix-rbm} provides proofs for results on RBM parameterizations
and data codings described in Section \ref{data-encoding}.

While applications of the RBM have claimed some successes in
classification problems, it is unclear if the model class allows one to
go beyond fitting to other statistical matters of importance in using
the models, such as quantification of the uncertainty of estimation and
the formulation of predictive distributions. These are closely tied to
the probability properties of the RBM model for explaining data
generation and the current exposition intends to contribute to a better
understanding in this direction.

@@

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}; R. Salakhutdinov and
Hinton \protect\hyperlink{ref-salakhutdinov2012efficient}{2012};
Srivastava, Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
stacking multiple layers of RBMs in a deep architecture, proponents of
the model are claimed to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(R. Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
is achieved by treating a hidden layer of one RBM as the visible layer
in a second RBM, and so on, until the desired multi-layer architecture
is created.

@@

The highly flexible nature of a RBM (having as it does
\(\nh + \nv + \nh*\nv\) parameters) creates at least three kinds of
potential issues of model impropriety that we will call
\emph{degeneracy}, \emph{instability}, and \emph{uninterpretability}. In
this section we define these, consider how to quantify them in a RBM,
and point out relationships among them.

@@

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
models, \(\mathcal{X}\) denotes all possible graphs that can be
constructed from a set of nodes and an exponentially parameterized
random graph model has a distribution with a probability mass function
of the form
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where \(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the
model parameter, and
\(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\) is a vector of
statistics based on the adjacency matrix of a graph. Here, as earlier,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
exponentially parametrized random graph model at \(\boldsymbol \theta\)
as \emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

@@

Considering exponential families of distributions, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) introduced
a concept of model deficiency related to \emph{instability}. Instability
can be roughly thought of as excessive sensitivity in the model, where
small changes in the components of potential data outcomes,
\(\boldsymbol x\), may lead to substantial changes in the probability
mass function \(f_{\boldsymbol \theta}(\boldsymbol x)\). Furthermore,
model instability can be viewed on a spectrum of potential sensitivity
in probability structure, with model degeneracy included as a extreme or
limiting case of instability. To quantify ``instability'' more
rigorously (particularly beyond the definition given by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011})) it is
useful to consider how RBM models might be expanded to incorporate more
and more visibles. When increasing the size of RBM models, it becomes
necessary to grow the number of model parameters (and in this process
one may also arbitrarily expand the number of hidden variables used). To
this end, let
\(\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1\),
denote an element of a sequence of RBM parameters indexed by the number
\(\nv\) of visibles \((V_1, \dots, V_\nv)\) and define a (scaled)
extremal log-probability ratio of the RBM model at
\(\boldsymbol \theta_{\nv}\) as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eq:elpr}
\end{align}
where
\(P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)\)
is the RBM probability of observing outcome \((v_1, \dots, v_\nv)\) for
the visible variables \((V_1, \dots, V_\nv)\) under parameter vector
\(\boldsymbol \theta_\nv\), after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles
(i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
(\eqref{eq:elpr}) needs to stay bounded for a sequence of RBM models to be
stable. That is, we make the following convention.
\BeginKnitrBlock{definition}[S-unstable RBM]
\protect\hypertarget{def:unnamed-chunk-1}{}{\label{def:unnamed-chunk-1}
\iffalse (S-unstable RBM) \fi{} }Let
\(\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh + \nh*\nv}, \nv \ge 1\),
be an element of a sequence of RBM parameters where the number of
hiddens, \(\nh \equiv \nh(\nv) \ge 1\), can be an arbitrary function of
the number \(\nv\) of visibles. A RBM model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, a RBM model sequence is
unstable if, given any \(c > 0\), there exists an integer \(n_c > 0\) so
that \(\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c\) for all
\(\nv \ge n_c\). This definition of \emph{S-unstable} is a
generalization or re-interpretation of the ``unstable'' concept of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) in that
here RBM models for visibles \((v_1, \dots, v_\nv)\) do not form an
exponential family and the dimensionality of \(\boldsymbol \theta_\nv\)
is not fixed, but rather grows with \(\nv\).

@@

In other words, if the probability ratio (\eqref{eq:elpr}) is too large,
then a RBM model sequence will exhibit large probability shifts for very
small changes in data configurations (i.e., will exhibit instability).
Such instability can be a concern in a model for reasons similar to
those related to degeneracy: as outcomes vary over the sample space, the
geography of probabilities is extremely rugged, with deep pits following
sharp mountains. Recall the applied example of RBM models as a means to
classify images. For data as pixels in an image, the instability result
in Proposition \ref{prp:instab1} manifests itself as a one pixel change
in an image (one component of the visible vector) resulting in a large
shift in the probability, which in turn could result in a vastly
different classification of the image. Examples of this behavior have
been presented in Szegedy et al.
(\protect\hyperlink{ref-szegedy2013intriguing}{2013}) for other deep
learning models, in which a one pixel change in a test image results in
a wildly different classification.

@@

We next explore and numerically explain the relationship between values
of \(\boldsymbol \theta\) and the three notions of model impropriety
(near-degeneracy, instability, and uninterpretability), for RBM models
of varying sizes.

@@

The results of this numerical study are summarized in Figures
\ref{fig:degen-plots}, \ref{fig:instab-plots}, and
\ref{fig:uninterp-plots}. From these three figures, it is clear that all
three measures of model impropriety show higher values for larger
magnitudes of the parameter vectors, supporting the RBM model properties
developed in Section \ref{background}. As a compounding issue, these
figures show that, as models grow in size, it becomes easier for more
parameter configurations to push RBM models into near-degeneracy,
instability and uninterpretability. Additionally, since there are
\(\nh*\nv\) interaction terms in \(\boldsymbol \theta\) versus only
\(\nh + \nv\) main effect terms, for large models there are many more
interaction parameters than main effects in the models. And so, severely
limiting the magnitude of the individual interactions may well help
prevent model impropriety. \par
\begin{figure}
\includegraphics{paper_files/figure-latex/degen-plots-1} \caption{Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size, where $\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})$ is split into main and interaction parameters. Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.}\label{fig:degen-plots}
\end{figure}
\begin{figure}
\includegraphics{paper_files/figure-latex/instab-plots-1} \caption{The sample mean value of $\text{ELPR}(\boldsymbol \theta)/V$ at each grid point for each combination of magnitude of $\boldsymbol \theta$ and model size. As the magnitude of $\boldsymbol \theta$ grows, so does the value of this metric, indicating typical instability in the model.}\label{fig:instab-plots}
\end{figure}
\begin{figure}
\includegraphics{paper_files/figure-latex/uninterp-plots-1} \caption{The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and the expectation vector given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$. Larger magnitudes of $\boldsymbol \theta$ correspond to larger differences, thus indicating reduced interpretability.}\label{fig:uninterp-plots}
\end{figure}
Figure \ref{fig:split-plots-dist} shows the fraction of near-degenerate
models for each magnitude of \(\boldsymbol \theta\) for each model
architecture. For each number \(\nv\) of visibles in the model, as the
number \(\nh\) of hiddens increase, the fraction near-degenerate
diverges from zero at increasing rates for larger values of
\(||\boldsymbol \theta||\). This shows that, as model size gets larger,
the risk of degeneracy starts at a slightly larger magnitude of
parameters, but very quickly increases until reaching close to 1. \par
\begin{figure}
\includegraphics{paper_files/figure-latex/split-plots-dist-1} \caption{The fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$. For each number $\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\boldsymbol \theta||$ as the number $\nh$ of hidden variables increases and the slope becomes steeper as $\nh$ increases as well.}\label{fig:split-plots-dist}
\end{figure}
These manageable examples indicate that RBMs are near-degenerate,
unstable, and uninterpretable for large portions of the parameter space
having large \(\|\boldsymbol \theta\|\). These problematic aspects
require serious consideration when using RBM models, on top of the
additional matter of principled/rigorous fitting of RBM models.

@@

Typically, fitting a RBM via maximum likelihood (ML) methods will be
infeasible due mainly to the intractability of the normalizing term
\(\gamma(\boldsymbol \theta)\) in a model (\eqref{eq:pmf}) of any
realistic size. Ad hoc methods are often recommended instead, which aim
to avoid this problem by using stochastic ML approximations that employ
a small number of MCMC draws (i.e., contrastive divergence, (Hinton
\protect\hyperlink{ref-hinton2002training}{2002})).

@@

appendix shows, not only is it possible to approximate any distribution
on the visibles arbitrarily well (cf. Montufar and Ay

@@

require at least \(4.7\) times as many iterations of the MCMC chain to
be run in order to achieve the same amount of effective information
about the posterior distribution. For this reason, consistent with the
ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an
effective method for fitting the RBM, though computing resources can
hinder use of the alternative BwTNML involving marginalization of hidden
variables.
